---
title: "Thema 09 Machine learning log"
author: "Dennis Haandrikman"
abstract: >
    During this experiment on machine learning, #TODO
date: "14-09-2022"
output:
    pdf_document:
        toc: true
        toc_depth: 3
---

```{r setup, include = FALSE}
# Knit options/settings
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)

# Required libraries
library("pander")
library("ggplot2")
```

# Introduction

## Importing the data
#### 15/09/2022, 16/09/2022, 20/09/2022

After acquiring the data from the database, import the data to see if there were any inconsistencies.

```{r import_data}
# Make sure the working directory is the main folder containing all the folders (project folder).
# Get the data files and import them.
bcel_data <- read.csv("./Data/input_bcell.csv", header = TRUE)
sars_data <- read.csv("./Data/input_sars.csv", header = TRUE)
covid_data <- read.csv("./Data/input_covid.csv", header = TRUE)

# print a quick head for the tables to see if it imported properly.
# Due to protein seq being such a lengthy tab, excluding from the head to make the head showable.
# In each dataset, the protein seq is the second column.
pander(head(bcel_data[-2]))
pander(head(sars_data[-2]))
pander(head(covid_data[-2]))
# here's the max length of a protein sequence to showcase why it was excluded from the head
max(nchar(bcel_data[2,]))
```

As we can see the information has been imported correctly.
However, it is of importance to check for any NA values in the information.
As such, an is.na() check will be performed on the data

```{r na_check}
# Using any(is.na(x)) showcases in true/false if there are any NA values.
any(is.na(bcel_data))
any(is.na(sars_data))
any(is.na(covid_data))
```
As we can see there aren't any missing values whatsoever in the datasets, as such we can continue to the actual EDA after we check the value-types each column has

```{r}
# Information for the bcel_data dataset
str(bcel_data)

# Information for the sars_data dataset
str(sars_data)

#Information for the covid_data dataset
str(covid_data)
```

We can see that the covid_data dataset doesn't contain the target column at the end of the dataset.

## EDA exploration
#### 20/09/2022


Now that we've made sure all the datasets have been imported properly and there are no missing values in the datasets.

Since the datasets all go over the properties of the peptide/proteins, a codebook was set up with information on all the properties.